Move Schedules: Fast persistence computations in coarse dynamic settings\thanks{\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}

### 2021-09-22 move_schedules.tex
Changes from jperea

@@

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title{Move Schedules: Fast persistence computations in sparse dynamic settings
	\thanks{\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}
}

% Authors: full names plus addresses.
\author{
	Matt Piekenbrock\thanks{Department of Computational Mathematics, Science \& Engineering, Michigan State University.}
	\and Jose A. Perea\thanks{Department of Mathematics, Michigan State University.}
}

@@

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. 
\begin{abstract}
	The standard procedure for computing the persistent homology of a filtered simplicial complex is the matrix reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles can be derived. 
	Persistence diagrams are known to vary continuously with respect to their input, which motivates the algorithmic study of persistence computations for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. In practice, the quadratic scaling in the number of  transpositions often makes this maintenance procedure slower than simply computing the decomposition from scratch, effectively limiting the application of dynamic persistence to relatively small data sets. In this work, we propose a coarser strategy for maintaining the decomposition over a discrete 1-parameter family of filtrations. Our first result is an analysis of a simple linear-time strategy which reduces the number of column operations needed to simulate persistence across a fixed homotopy by at most a factor of 2. We then show a modification of this technique which maintains only a sublinear number of valid states, as opposed to a quadratic number of states, and we provide tight lower bounds for this technique.
	Finally, we provide empirical results suggesting that the decrease in operations needed to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states, and the proposed sublinear coarsening.
	Applications to multi-dimenisonal persistence and crocker stacks are also presented. 
\end{abstract}

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a sufficiently tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that small changes in the underlying function produce small changes in the corresponding diagrams:  if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the study and application of persistence to time-varying or dynamic settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity (e.g. essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}), their adoption to dynamic settings poses a challenging computational problem.
With the current state of the art tools, there is no recourse to the problem of computing persistence from a time-varying filtration containing millions of simplices at thousands of snapshots in time.
An efficient means of updating the persistence of a [time-varying] filtration has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become computationally viable tools in dynamic  settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation given a time-varying filtration. 
The vineyards approach can be interpreted as an extension of the \emph{matrix reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time.
The main step in the reduction algorithm is a specific decomposition $R = D V$ of the boundary matrix $D$ of $K$. 
Since $V$ is always full rank (and upper-triangular), an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm transforms a time-varying filtration into an ordered set of adjacent transpositions to be applied to the rows and columns of the decomposition $D = RU$, each of which take at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for  homological critical points can only occur when the filtration order changes. 
This algorithm is also efficient  asymptotically: if there are $d$ time-points where this filtration order changes, then the vineyards algorithm takes $O(m^3 + md)$ time: one $O(m^3)$-time reduction upfront for the filtration at time $t_0$ followed by one $O(m)$ operation to update the decomposition for the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams independently at every time point.

Despite its theoretical efficiency, vineyards is often not the method of choice in practical settings. 
Indeed, while there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc... see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 has an implementation of vineyards, but the algorithm was never ported to the 2nd version.} The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' Indeed, whether performing the reductions naively or updating the decomposition dynamically using vineyards, they observed that obtaining the decomposition is the most computationally demanding aspect of their 2-D persistence algorithm.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards are  efficient for   \emph{continuous} time families of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them are sufficient.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in in~\cite{busaryev2010tracking}, we are able to make a tradeoff in the number of times the decomposition is restored to a valid state, dramatically reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse contexts, making dynamic persistence more computationally tractable for a wider class of use-cases.

@@

 \subsection{A Motivating Example and Contributions}\label{sec:motivation} 
As a motivating example, we describe a simple experiment illustrating why the vineyards algorithm does not always yield an efficient method for time-varying applications. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, where scale of the annulus is parameterized by a single parameter $\tau$. Thus, the data set is a 1-parameter family of images. Each image contains the same number of pixels whose intensities---these are determined by $\tau$---vary with time. For each image, we build a filtered simplicial complex using the \emph{Freudenthal} triangulation, with the ordering of inclusions obtained by lower stars.  
 \begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.36\textwidth, height = 1.45in]{circle_vineyards.png}
 	\includegraphics[width=0.63\textwidth, height = 1.45in]{circle_complex.png}
 	\caption{On the right, the grayscale image data set and corresponding lower star filtrations. On the left, the cumulative number of column operations required to compute persistence for this time-varying filtration. The persistent Betti numbers for each filtration from left to right are: $\beta_1 = (1,0,0), \beta_2 = (1,1,0), \beta_3 = (1,1,0), \beta_4 = (1,1,0), \beta_5 = (4,0,0)$. Observe computing $10$ diagrams independently at evenly spaced time points (the green line) both captures these major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}
Snapshots of the image data and its corresponding time-varying filtration is depicted in Figure~\ref{fig:vineyards} (top right and bottom right, respectively), which we use as a baseline for comparing the standard algorithm $\texttt{pHcol}$ with vineyards. Note the size of the filtration is fixed--no simplices are ever added or removed over time. 
There are two events which dramatically alter the persistence diagrams extracted from these filtrations: the first occurs when the central connected component splits to form a cycle, and the second occurs when the annulus approaches the outer dimensions of the grid, splitting into four components. As a consequence, in this example, only a few persistence diagrams are needed to capture the major topological changes over time.

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches which one can use to simulate persistence dynamically. 
  Since it is unknown a priori at which times  the persistent homology will change, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagrams at each one of them  independently. An alternative approach is to construct a homotopy between two time-points, and then decompose this homotopy into adjacent transpositions--the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. As one can see from the figure, if one needs to simulate the homotopy exactly, the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is sparse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

  The existence of a time discretization that is more efficient to compute than continually updating the decomposition indicates that the vineyards algorithm must incur some amount of overhead (in terms of column operations) to perform the transpositions. That is, even in situations where one is updating a decomposition to a filtration that is  relatively ``close'', e.g. any two adjacent filtrations in the bottom right of Figure~\ref{fig:vineyards}, it is still more efficient to apply \texttt{pHcol} directly than  to iteratively update the decomposition (see the case where $n = 10$). 
 Indeed, any optimizations to the reduction algorithm (for example, the clearing optimization~\cite{chen2011persistent}) would only increase this disparity, making the vineyards algorithm less computationally attractive even in settings where $n$ is larger. Since dynamic or time-varying are common settings where persistence is applied (see ~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal}), it is worth studying this phenomenon.

\textbf{Our results and approach are as follows:} First, we leverage work of Busaryev et. al.~\cite{busaryev2010tracking} to include a coarser type of operation to the dynamic setting, called \emph{moves}. We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case bounds in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$, which we call \emph{schedules}. 
We investigate the feasibility of optimizing the cost 
(i.e., number of column operations) of these schedules directly, including the existence of a greedy-type approach, which we show can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective to minimize, provide bounds justifying its relevance to the original objective, and give an efficient $O(d \log d)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistent homology computations is given, wherein our approach is demonstrated to be an order of magnitude more efficient than existing approaches in applications such as flock simulation and 2D persistence computation. We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

In contrast, there is extensive work  on improving the efficiency of computing a (static) $R = DV$ decomposition. Chen~\cite{chen2011persistent} proposed \emph{persistence with a twist}, also called the \emph{clearing optimization}, which exploits a boundary/cycle relationship to ``kill'' columns early in the reduction rather than reducing them. 
Another popular optimization is to utilize the duality between   homology and cohomology~\cite{de2011dualities}, which dramatically improves the effectiveness of the clearing optimization~\cite{bauer2021ripser}. 
There are many other optimizations on the implementation side: the use of ranking functions defined on the combinatorial number system enables implicit cofacet enumeration, removing the need to store the boundary matrix explciitly; the apparent/emergent pairs optimization identifies columns whose pivot entries are unaffected by the reduction algorithm, reducing the total number of columns which need be reduced; sparse data structures such as bit-trees and lazy heaps allow for efficient column-wise additions with $\mathbb{Z}_2 = \mathbb{Z}/2\mathbb{Z}$ coefficients and effective $O(1)$ pivot entry retrieval, and so on~\cite{bauer2021ripser, bauer2017phat}.

@@

If  zeroth homology is the only dimension of interest, computing and updating both the persistence and rank information of the persistent homology groups is greatly simplified. For example, if the relations are available a-priori, obtaining a tree representation fully characterizing the connectivity of the underlying space (also known as the \emph{incremental connectivity} problem) takes just $O(\alpha(n) n)$ time using the disjoint-set data structure, where $\alpha(n)$ is the extremely slow-growing inverse Ackermann function. 
Adapting this approach to the time-varying setting, Oesterling et al.~\cite{oesterling2015computing} give a $O(e)$-per-update algorithm that allows one to maintain a \emph{merge tree} with $e$ edges,
associated to the filtration parameter changing continuously over time. 
If only the zeroth-dimensional Betti numbers are needed for a particular application, this problem reduces even further to tracking the connected components of a dynamic graph---sometimes referred to as the \emph{dynamic connectivity problem}, which can be efficiently  solved in amortized $O(\log n)$ query and update times using either Link-cut trees or multi-level Euler tour trees~\cite{kapron2013dynamic}.

\subsection{Outline} The remainder of the paper is organized as follows: we review and establish the notation we will use to describe simplicial complexes, persistent homology, and dynamic persistence. We also cover the reduction algorithm (designated here as \texttt{pHcol}), the vineyards algorithm, and the set of \emph{move}-related algorithms introduced in~\cite{busaryev2010tracking}, which serves as the starting point of this work. 
In section 3 we introduce our proposed alternative to vineyards and provide efficient algorithms to make this alternative viable. 
In section 4 we demonstrate a few applications of the proposed method. In section 5 we conclude the paper by discussing other possible applications and future work. 
We also include an appendix which introduces certain topics in more detail, such as the vineyards algorithm itself, for readers who may be unfamiliar but would like a more in-depth understanding.

@@

Suppose one has a family $\{K_i\}_{i\in I}$ of simplicial complexes indexed by a totally ordered index set $I$, and so that for any $i< j \in I$ we have $K_i \subseteq K_j$. There are two index sets of interest in this effort: $\mathbb{R}$ and $[n] = \{ 1, \dots, n\}$. 
Such a family is called a \emph{filtration},
which is deemed to be \emph{essential} if $i \neq j$ implies $K_i \neq K_j$.
Moreover, an essential filtration is said to be  \emph{simplexwise} if $K_j \smallsetminus K_i = \{\sigma_j\}$ for all $i< j \in I$ with $K_j \neq \emptyset$.
Any filtration may be trivially converted into an essential, simplexwise filtration via a set of \emph{condensing}, \emph{refining}, and \emph{reindexing} maps (see~\cite{bauer2021ripser} for more details). 
As a result, and without loss of generality, here we exclusively consider essential simplexwise filtrations. For brevity, we will simply refer to them as filtrations.

>>>>

\newsiamremark{remark}{Remark}

@@

% Title. If the supplement option is on, then "Supplementary Material"
% is automatically inserted before the title.
\title{Move Schedules: Fast persistence computations in coarse dynamic settings\thanks{\textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}
}

% Authors: full names plus addresses.
\author{
	Matt Piekenbrock\thanks{Department of Computational Mathematics, Science \& Engineering, Michigan State University.}
	\and Jose A. Perea\thanks{
	Department of Computational Mathematics, Science \& Engineering, 
	Department of Mathematics, Michigan State University.}
}

@@

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. 
\begin{abstract}
	The standard procedure for computing the persistent homology of a filtered simplicial complex is the matrix reduction algorithm. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles can be derived. 
	Persistence diagrams are known to vary continuously with respect to their input, which motivates the algorithmic study of persistence computations for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	In practice, the quadratic scaling in the number of  such transpositions often makes this maintenance procedure slower than simply computing the decomposition from scratch,  limiting the application of dynamic persistence to relatively small data sets. In this work, we propose a coarser strategy for maintaining the decomposition over a discrete 1-parameter family of filtrations. Our first result is an analysis of a simple linear-time strategy which reduces the number of column operations needed to simulate persistence across a fixed homotopy by at most a factor of 2. We then show a modification of this technique which maintains only a sublinear number of valid states, as opposed to a quadratic number of states, and we provide tight lower bounds for this technique.
	Finally, we present empirical results suggesting that the decrease in operations needed to compute diagrams across a family of filtrations is proportional to the difference between the expected quadratic number of states, and the proposed sublinear coarsening.
	Applications to multi-dimenisonal persistence and crocker stacks are also presented. 
\end{abstract}

@@

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a sufficiently tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that small changes in the underlying function produce small changes in the corresponding diagrams:  if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the study and application of persistence to time-varying or dynamic settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
With the current state of the art tools, there is no recourse to the problem of computing persistence from a time-varying filtration containing millions of simplices at thousands of snapshots in time.
Efficiently  updating the persistence of a time-varying filtration has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become computationally viable tools in dynamic  settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation given a time-varying filtration. 
The vineyards approach can be interpreted as an extension of the \emph{matrix reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time.
The main step in the reduction algorithm is a specific decomposition $R = D V$ of the boundary matrix $D$ of $K$. 
Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm transforms a time-varying filtration into an ordered set of permutations to be applied to adjacent rows and columns of the decomposition $D = RU$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for  homological critical points can only occur when the filtration order changes. 
This algorithm is also efficient  asymptotically: if there are $d$ time-points where the filtration order changes, then   vineyards   takes $O(m^3 + md)$ time; one $O(m^3)$-time reduction upfront for the filtration at time $t_0$ followed by one $O(m)$ operation to update the decomposition for the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams independently at every time point.

Despite its theoretical efficiency, vineyards is often not the method of choice in practical settings. 
Indeed, while there is an increasingly rich ecosystem of software packages offering variations of the standard reduction algorithm (e.g. Ripser, PHAT, Dionysus, etc... see~\cite{otter2017roadmap} for an overview), implementations of the vineyards algorithm are relatively uncommon.\footnote{Dionysus 1 has an implementation of vineyards, but the algorithm was never ported to the 2nd version.} 
The reason for this disparity is perhaps explained by Lesnick and Wright~\cite{lesnick2015interactive}: ``While an update to an $RU$ decomposition involving few transpositions is very fast in practice... many transpositions can be quite slow... it is sometimes much faster to simply recompute the $RU$-decomposition from scratch using the standard persistence algorithm.'' 
Whether performing the reductions naively or updating the decomposition dynamically using vineyards, they observe that this is the most computationally demanding aspect of  RIVET \cite{rivet}.

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling   \emph{continuous} time families of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them are sufficient.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in in~\cite{busaryev2010tracking}, we are able to make a tradeoff in the number of times the decomposition is restored to a valid state, dramatically reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse contexts, making dynamic persistence more computationally tractable for a wider class of use-cases.

@@

 \subsection{A Motivating Example and Contributions}\label{sec:motivation} 
As a motivating example, we describe a simple experiment illustrating why the vineyards algorithm does not always yield an efficient method for time-varying applications. 
Consider a series of grayscale images (i.e., a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, where the scale of the annulus is parameterized by a single parameter $\tau$. 
Each image contains the same number of pixels whose intensities---these are determined by $\tau$---vary with time. For each image, we build a filtered simplicial complex using the \emph{Freudenthal} triangulation of the plane, with the order of inclusions obtained by lower stars.  
 \begin{figure*}[t]
 	\centering
 	\includegraphics[width=0.36\textwidth, height = 1.45in]{circle_vineyards.png}
 	\includegraphics[width=0.63\textwidth, height = 1.45in]{circle_complex.png}
 	\caption{On the right, the grayscale image data set and corresponding lower star filtrations. On the left, the cumulative number of column operations required to compute persistence for this time-varying filtration. The   Betti numbers for each of the shown  steps in the filtration, from left to right, are: $(\beta_0,\beta_1, \beta_2)  = (1,0,0), \; (1,1,0),  \; (1,1,0), \; (1,1,0), \; (4,0,0)$. Observe that computing the diagrams independently at 10 evenly spaced time points (the green line) both captures these major topological changes and is the most computationally efficient approach shown.}
 	\label{fig:vineyards}
 \end{figure*}
Snapshots of the image data and its corresponding time-varying filtration is depicted in Figure~\ref{fig:vineyards} (top right and bottom right, respectively), which we use as a baseline for comparing the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}) with vineyards. Note that the size of the filtration is fixed--no simplices are ever added or removed over time. 
There are two events which dramatically alter the persistence diagrams extracted from these filtrations: the first occurs when the central connected component splits to form a cycle, and the second occurs when the annulus approaches the outer dimensions of the grid, splitting into four components. As a consequence, in this example, only a few persistence diagrams are needed to capture the major topological changes over time.

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches.
  Since it is unknown a priori at which times  the persistent homology will change, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagrams at each one of them  independently. An alternative approach is to construct a homotopy between two contiguous time-points, and then decompose this homotopy into adjacent transpositions--the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. 
  The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. 
  As one can see from the figure, if one needs to simulate the homotopy exactly, then the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is sparse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

  The existence of a time discretization that is more efficient to compute than continually updating the decomposition, indicates that the vineyards algorithm must incur some amount of overhead (in terms of column operations) to perform transpositions, even if the persistence diagrams remain unchanged. 
  Even in situations where one is updating a decomposition to a filtration that is  relatively ``close'', e.g. any two adjacent filtrations in the bottom right of Figure~\ref{fig:vineyards}, it is still more efficient to apply \texttt{pHcol} directly than  to iteratively update the decomposition (see the case where $n = 10$). 
 Moreover, any optimization to the reduction algorithm (e.g.,  clearing~\cite{chen2011persistent}) would only increase this disparity.
 See~\cite{topaz2015topological, xian2020capturing, lesnick2015interactive, kim2020spatiotemporal} 
 for relevant scenarios and applications.

\textbf{Our approach and contributions are as follows:} First, we leverage the \emph{moves} framework of Busaryev et. al.~\cite{busaryev2010tracking} to include  coarser  operations---in terms of valid intermediate decomposition states, as compared to vineyards---for the dynamic setting. 
We give a tight lower bound on the number of moves needed to perform an arbitrary permutation to the $R = D V$ decomposition, and give a proof of optimality by a reduction to the permutation edit distance problem.
We also give worst-case bounds in expectation as well as efficient algorithms for achieving these bounds---both of which are derived from a reduction to the Longest Increasing Subsequence (LIS) problem. 
This reduction yields an efficient algorithm for generating sequences of moves $( s_1, s_2, \dots, s_d )$ of minimal size $d$, which we call \emph{schedules}. 
However, not all valid schedules incur the same permutation costs 
(i.e., number of column operations).
We investigate the feasibility of choosing optimal cost schedules, and show that 
greedy-type approaches can lead to arbitrarily bad behavior. 
In light of these results, we give an alternative proxy-objective for cost minimization, provide bounds justifying its relevance to the original problem, and give an efficient $O(d \log d)$ algorithm for approximately solving this proxy minimization. 
A performance comparison with other reduction-based persistent homology computations is given; 
our move schedules are demonstrated to be an order of magnitude more efficient than existing approaches in applications such as flock simulation and 2D persistence computations. 
%We conclude with a discussion of situations where our strategy is most applicable and future directions for research.

@@

In contrast, there is extensive work  on improving the efficiency of computing a (static) $R = DV$ decomposition. Chen~\cite{chen2011persistent} proposed \emph{persistence with a twist}, also called the \emph{clearing optimization}, which exploits a boundary/cycle relationship to ``kill'' columns early in the reduction rather than reducing them. 
Another popular optimization is to utilize the duality between   homology and cohomology~\cite{de2011dualities}, which dramatically improves the effectiveness of the clearing optimization~\cite{bauer2021ripser}. 
There are many other optimizations on the implementation side: the use of ranking functions defined on the combinatorial number system enables implicit cofacet enumeration, removing the need to store the boundary matrix explicitly; the apparent/emergent pairs optimization identifies columns whose pivot entries are unaffected by the reduction algorithm, reducing the total number of columns which need be reduced; sparse data structures such as bit-trees and lazy heaps allow for efficient column-wise additions with $\mathbb{Z}_2 = \mathbb{Z}/2\mathbb{Z}$ coefficients and effective $O(1)$ pivot entry retrieval, and so on~\cite{bauer2021ripser, bauer2017phat}.

@@

If  zeroth homology is the only dimension of interest, computing and updating both the persistence and rank information  is greatly simplified. For example, if the relations are available a-priori, obtaining a tree representation fully characterizing the connectivity of the underlying space (also known as the \emph{incremental connectivity} problem) takes just $O(\alpha(n) n)$ time using the disjoint-set data structure, where $\alpha(n)$ is the extremely slow-growing inverse Ackermann function. 
Adapting this approach to the time-varying setting, Oesterling et al.~\cite{oesterling2015computing} give a $O(e)$-per-update algorithm that allows one to maintain a \emph{merge tree} with $e$ edges,
associated to the filtration parameter changing continuously over time. 
If only the zeroth-dimensional Betti numbers are needed for a particular application, this problem reduces even further to tracking the connected components of a dynamic graph---sometimes referred to as the \emph{dynamic connectivity problem}, which can be efficiently  solved in amortized $O(\log n)$ query and update times using either Link-cut trees or multi-level Euler tour trees~\cite{kapron2013dynamic}.

\subsection{Outline} The remainder of the paper is organized as follows: we review and establish the notations we will use to describe simplicial complexes, persistent homology, and dynamic persistence in Section \ref{sec:background}. 
We also cover the reduction algorithm (designated here as \texttt{pHcol}), the vineyards algorithm, and the set of \emph{move}-related algorithms introduced in~\cite{busaryev2010tracking}, which serves as the starting point of this work. 
In Section \ref{sec:move_schedules} we introduce our move schedules  and provide efficient algorithms to make this alternative viable. 
In Section \ref{sec:results} we demonstrate a few applications of the proposed method. 
In particular, to the computation of crocker stacks from flock simulations, and to 2-dimensional persistence on the Mumford data set of image patches from natural images. 
In Section \ref{sec:conclusion} we conclude the paper by discussing other possible applications and future work. 
We also include an appendix which introduces certain topics in more detail, such as the vineyards algorithm itself, for readers who may be unfamiliar but would like a more in-depth understanding.

@@

Suppose one has a family $\{K_i\}_{i\in I}$ of simplicial complexes indexed by a totally ordered   set $I$, and so that for any $i< j \in I$ we have $K_i \subseteq K_j$. There are two index sets of interest in what follows: $\mathbb{R}$ and $[n] = \{ 1, \dots, n\}$. 
Such a family is called a \emph{filtration},
which is deemed to be \emph{essential} if $i \neq j$ implies $K_i \neq K_j$.
Moreover, an essential filtration is said to be  \emph{simplexwise} if $K_j \smallsetminus K_i = \{\sigma_j\}$ 
whenever $j$ is the immediate successor of $i$ in $I$.
Any finite filtration may be trivially converted into an essential, simplexwise filtration via a set of \emph{condensing}, \emph{refining}, and \emph{reindexing} maps (see~\cite{bauer2021ripser} for more details). 
As a result, and without loss of generality, here we exclusively consider essential simplexwise filtrations. For brevity, we will simply refer to them as filtrations.

Let $K$ be an abstract simplicial complex and $\mathbb{F}$ a field.
A $p$-chain is a
formal $\mathbb{F}$-linear combination  of $p$-simplices of $K$. The collection of $p$-chains under addition yields an 
$\mathbb{F}$-vector space  denoted   $C_p(K)$. 
The $p$-boundary $\partial_p(\sigma)$  of a $p$-simplex $\sigma\in K$ is the alternating sum of its oriented co-dimension 1 faces; the $p$-boundary of a $p$-chain is defined 
linearly in terms of its constitutive simplices. 
A $p$-chain with zero boundary is called a $p$-cycle, and together they form $Z_p(K) = \mathrm{Ker}\,\partial_p$. 
Similarly, the collection of $p$-boundaries forms  $B_p(K) = \mathrm{Im}\,\partial_{p+1}$.
Since $\partial_p \circ \partial_{p+1} = 0$ for all $p\geq 0$, 
then the quotient space $H_p(K) = Z_p(K) / B_{p}(K)$ is well-defined, and called the 
$p$-th homology of $K$ with coefficients in $\mathbb{F}$. 
If $\{K_i\}_{i\in [m]}$ is a filtration, then the inclusion maps  $K_i\subset K_j$  for $i < j$, induce linear transformations $f_p^{i,j}: H_p(K_i) \to H_p(K_j)$ at the level of homology:
\begin{equation}
	H_p(K_1) \to H_p(K_2) \to \dots \to H_p(K_m)
The $p$-th persistent homology groups are the images of these transformations: $H_{p}^{i,j} = \mathrm{Im}\,f_p^{i,j}$. 
%Note that if $i = j$, then $H_{p}^{i,j} = H_{p}(K_i) = H_{p}(K_i)$ is   just the ``standard'' homology. 
Simplices whose inclusion in the filtration creates a new homology class are   called \emph{creators}, and simplices that destroy homology classes are   called \emph{destroyers}. 
The filtration indices of 
these creators/destroyers are referred to as \emph{birth} and \emph{death} times, respectively. 
The collection of birth/death  pairs 
$(i,j)$ is denoted $\mathrm{dgm}_p(K)$, 
and referred to as the $p$-th \emph{persistence diagram} of $K$.
If a homology class is born at time $i$ and dies entering time $j$, the difference $\lvert i - j \rvert$ is called the \emph{persistence} of that class.
In practice, filtrations often arise from triangulations parameterized by geometric scaling parameters, and the ``persistence'' of a homology class actually refers to its lifetime with respect to the scaling parameter. 
%For this reason, we'll refer to the difference $\lvert i - j \rvert$ of a homology class $[c]$ as its \emph{index persistence}.
\noindent \textbf{Example 2.1:} Non-essential filtrations often arise in geometrical contexts. For example, given a finite metric space $(X, d)$, the \emph{Vietoris-Rips} complex at scale $\epsilon \in \mathbb{R}$ is the abstract simplicial complex given by: 
\begin{equation}\label{eq:rips}
	\mathrm{Rips_\epsilon}(X) = \{ S \subseteq X : S \neq \emptyset \;\mbox{ and }\;\mathrm{diam}(S) \leq \epsilon \}
The corresponding filtration is called the \emph{Vietoris-Rips} (VR) filtration, indexed by $I = \mathbb{R}$. 
A VR filtration indexed over $\mathbb{R}$  can be condensed to an essential filtration indexed over the set of pairwise distances $\{\, d(x,x') \; | \; x,x' \in X \, \}$, 
which can then be further reindexed to an essential simplexwise filtration by extending a total order of $X$ to its power set---e.g., 
using the lexicographical ordering. 
